---
title: 'ASM TS Project: RenewUSA'
author: "Kathryn Weissman & Carol Azparrent"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
rm(list=ls())
#setwd("~/Documents/ASM/Project")
```

Total production of renewable energy in USA (trillions of BTU-British Thermal Units)

https://www.eia.gov/totalenergy/data/browser/index.php?tbl=T10.01#/?f=M&start=199001&end=201901&charted=6-7-8-9-14

According to the source data website, this time series is an aggregation of different renewable energy sources including: hydroelectric power, geothermal, solar, wind, and biomass.

Biomass, wind, and solar energy have become more popular starting at different times after 2005 which explains some of the variation in the series.


```{r renewUSA}
series = window(ts(read.table("data/RenewUSA.dat", header=F),start=1990,freq=12))
series
```

```{r plotSeries}
plot(series,main="Total production of renewable energy (USA)", ylab='Trillion Btu')
abline(v=1990:2020,col=4,lty=3)
```
1. Identification
1.a) Transform into stationary Time Series

## is the variance constant?
```{r}
boxplot(series~floor(time(series))) 
```
```{r}
#boxplot for periods of 12 observations = 1 year
groupedserie <- matrix(series[1:(20*12)], ncol = 20) #boxplot of each year
boxplot(groupedserie)
```
```{r}
# groupedserie : serie with periods (periods of 12 months = 1 year)
m<-apply(groupedserie,2,mean)
v<-apply(groupedserie,2,var)
plot(v~m)
abline(lm(v~m),col=2,lty=3) #col: color of line, lty:type of line
# for mean vs variance (v = s^2) -> there is an increasing
# ===> THE VARIANCE IS NOT CONSTANT --> we have to take logarithmic
```
```{r}
groupedserie
```

Variance not constant: change the scale = logarithms
```{r}
# now, we apply the logarithmic to "series"
lnserie=log(series)
plot(lnserie)
abline(v=1990:2020,col=4,lty=3)
abline(h=0)
```


## is there a seasonal pattern?

```{r}
monthplot(lnserie)
#there is a seasonal pattern
```
```{r}
ts.plot(matrix(lnserie,nrow=12),col=1:8)
#most of them are in parallel -> so there is a seasonal pattern
```
so, we take seasonal difference
```{r}
d12lnserie <- diff(lnserie, lag=12)
plot(d12lnserie)
#12 because it's 12 months data in a year
```
```{r}
monthplot(d12lnserie) #no more seasonal pattern
```
```{r}
ts.plot(matrix(d12lnserie,nrow=12),col=1:8)
#no more seasonal pattern (not in parallel)
```
now we work with "d12lnserie"


## is the mean constant?
```{r}
plot(d12lnserie)
abline(h=0)
abline(h=mean(d12lnserie),col=2)
#constant mean = yes
# this mean in 0? no
```
```{r}
mean(d12lnserie)
```
add regular difference:
```{r}
d1d12lnserie = diff(d12lnserie)
plot(d1d12lnserie)
abline(h=0)
abline(h=mean(d1d12lnserie), col=2)
```
```{r}
mean(d1d12lnserie) 
#mean different from 0  --> differentation
```

Comparing the variance to choose the best one:
```{r}
var(d12lnserie) #d=0
var(d1d12lnserie) #d=1 **************
var(diff(d1d12lnserie)) #d=2

```
d=1 wins as it has the lower variance. 
We would take only one differentation.
W_t = (1-B)(1-B^12)X_t , d=1

we applied first log X_t then, seasonal difference (s=12) as we had seasonal pattern, and finally we added one regular difference.

So now, we work with "d1d12lnserie"
we have:
- seasonal difference
- one regular difference



1.b) ################# ACF/PACF regular & seasonal #####################################
NEXT PART: Analysis ACF/PACF  W_t ==>> p,q, P,Q

seasonal part: (P,D,Q)_12
regular part: p,d,q
ARMA(p,d,q)(P,D,Q)_12
```{r}
# now we work only with d1d12lnserie
par(mfrow=c(1,2))
acf(d1d12lnserie, ylim=c(-1,1), lag.max = 72,col=c(2,rep(1,11)),lwd=2)
pacf(d1d12lnserie, ylim=c(-1,1), lag.max = 72,col=c(rep(1,11),2),lwd=2)
par(mfrow=c(1,1))

#SEASONAL: red lines
#REGULAR: black lines
```
